Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-04 analysis/paper/paper.Rmd
updates for introduction and add references

The development of craft specialization is usually associated with the social changes since it reflects production systems that inform social, economic, or political conditions of a society. Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp. 43]. The implication of the emergence of specialization for social structure or social process has been a theme in archaeological investigations. According to the political approaches, craft specialization, attached production specifically, is viewed as a strategy used by elites to control over production systems for the sake of power and authority [@Costin2001]. The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999]. Economic models consider production, exchange, and consumption as a whole and explain that specialization is a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], or a reaction to the changes in subsistence practices [@Stark1995economic]. However, the explanation for the emergence of specialized production might not be exclusive or straightforward. Furthermore, specialization is not necessarily accompanied with social differentiation observed in lowland northern Philippines with a low degree of ceramic specialization in Ifugao societies with social ranking during the Spanish colonization [@Acabado2018]. This implies that the condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2015].

## social changes in cross-cultural interaction contexts
- here I have the third cultural resistance, identity take a look at social boundary
- hypothesis, goals
- trade [@Alizadeh2018]

Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies. We explore the culture changes in indigenous societies in northeastern Taiwan that experienced foreign interactions with the Europeans and Chinese since 17th century by examining the local pottery production. We observed ceramic production in northeastern Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization. Our aim is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes. Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels? How does pottery change in shape throughout 600 years at this site? Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?

Our hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals. The first step is to detecting pottery standardization when discussing the emergence of craft specialization and the possible changes in social organization. This paper uses Elliptic Fourier Analysis, one outline approach in Geometric morphometric methods, to test whether there was increasing pottery standardization in relation to foreign contacts.

## Measuring ceramic standardization

The variation of standardization of specialized product is commonly used as indicator to identify craft specialization based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation between sets of data coupled with statistical comparison that lower value reflects higher degree of standardization [@Eerkens2001; @Roux2003; @Roux2018; @Stark1995].  However, this traditional typological and linear measurements might be limited because they can be insensitive to subtle variations resulting from changes in assembles. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Taking geometric morphomerics approaches to the measurement of craft specialization, we study shape and standardization of locally made ceramic at Kiwulan (1350-1950 AD), a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts. The aim of this paper is to investigate if there are increasing ceracmic specialization resulting from interaction with the Europeans and the Chinese, two major foreign influences in early historical Taiwan that might indicate social changes in the indigenous society.

@@

However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks convey important information. In addition, obvious and reliable landmark points may be unavailable for complex or rounded shaped structures. In those cases, outline approaches serve as a solution by analyzing the overall shape of a morphological structure. One of the outline approaches is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013]. Another approach is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape which turns coordinates into coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adams2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometry of an object. The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for the desired number of harmonics [@Bonhomme2014]. The application of EFA in archaeology is mostly used for bone morphology to understand the biological variation and recently more for stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Outline approach based on EFA is suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variations, for example, ceramics. @Wilczek2014 evaluate the concordance between outline-based approaches, the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. They found that the variation in ceramic shapes analyzed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes. In addition, the results of shape variation by EFA helps to understand the level of production standardization over time across that region.

@@

# Materials

## Archaeological background

@@

- identity, social boundary

The results confirm that the differences in pottery shapes can be detected using outline analysis approaches. It shows there is differences in pottery shapes between pre-European and post-European, and Pre-European and Chinese contact. A increasing standardization of shape of pottery during the Chinese contact period compared to the other two periods is observed. This study provide an comparable basis for our understanding of the standardized pottery as an indicator of social changes in a cultural contact conditions. Northeastern Taiwan had been encountered two major foreign contacts in the 17th century with the Europeans and 19th century with the Chines immigrants. It shows that the standardization of pottery shape appears during the European presence and increased in the Chinese period. This may suggest that an emergence of specialized groups or a consensus of the adoption of similar shape of pottery in the society. The changes in the standardization of pottery may indicate a major social changes in the local indigenous organization. When it comes to ceramic specialization, metric measurements and clay paste are two major indicators. Shape analysis could be viewed as a replacement of metric measurement that enables subtle changes to be detected. We observe the increasing standardization of the pottery shape, however, it is interesting to note that there is no obvious changes in the clay paste across different phases.

>>>>

The development of craft specialization is usually associated with the social changes since it reflects production systems that inform social, economic, or political conditions of a society. Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp. 43]. The implication of the emergence of specialization for social structure or social process has been a theme in archaeological investigations. According to the political approaches, craft specialization, attached production specifically, is viewed as a strategy used by elites to control over production systems for the sake of power and authority [@Costin2001]. The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999]. Economic models consider production, exchange, and consumption as a whole and explain that specialization is a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], or a reaction to the changes in subsistence practices [@Stark1995economic]. However, the explanation for the emergence of specialized production might not be exclusive or straightforward. Furthermore, specialization is not necessarily accompanied with social differentiation observed in lowland northern Philippines with a low degree of ceramic specialization in Ifugao societies with social ranking during the Spanish colonization [@Acabado2018]. This implies that the condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2015; @Costin2001].

The emergence of social inequality in the context of cross-cultural interaction especially in a colonial situation was observed in many parts of world when the introduction of foreign trade goods to local indigenous societies. The monopolization of long-distance trade goods might cause substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Recently more studies discuss the indirect effects of colonialism or pericolonial archaeology that investigates areas where European colonial rule was limited or their conquests were unsuccessful, but their colonial activities had economically and politically impacts on indigenous peoples in the periphery of colonial control [@Trabert2017; @Acabado2017]. For example, @Acabado2017's study on Ifugao society in the highland Philippines suggests economic and political intensification during the Spanish presence that indicates the responses of indigenous peoples to Spanish cooptation. The archaeological evidence in northeastern Taiwan show evidence of use of prestige goods such as trade ornaments ornaments and their uneven distribution pattern in domestic area and burial contexts indicates some extent of social inequality when encountered by the Spanish and the Dutch in the 17th century. In addition to the differential distribution of trade ornaments, the locally made ceramic present a high consistency in form and shape compared to other pottery found from other regions on this island that hints the possibility of pottery specialization. To further understand the relationship among the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers, this paper uses pottery production as a proxy for identifying social changes in a cross-cultural contact context.

To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation between sets of data coupled with statistical comparison that lower value reflects higher degree of standardization [@Eerkens2001; @Roux2003; @Roux2018; @Stark1995]. However, this traditional typological and linear measurements might be limited because they can be insensitive to subtle variations resulting from changes in assembles. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Taking geometric morphomerics approaches to the measurement of craft specialization, we study shape and standardization of locally made ceramic at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts. The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans and the Chinese, two major foreign influences in early historical Taiwan that might indicate social changes in the indigenous society.

Our hypothesis is that if foreign contacts including European in the 17th century and Chinese in the 19th century had impacts on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals. The aim of this paper is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes. Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels? How does pottery change in shape throughout 500 years at this site? Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?

@@

However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks convey important information. In addition, obvious and reliable landmark points may be unavailable for complex or rounded shaped structures. In those cases, outline approaches serve as a solution by analyzing the overall shape of a morphological structure. One of the outline approaches is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013]. Another approach is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape which turns coordinates into coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adams2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometry of an object. The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for the desired number of harmonics [@Bonhomme2014]. The application of EFA in archaeology is mostly used for bone morphology to understand the biological variation and recently more for stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].

Outline approach based on EFA is suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variations, for example, ceramics. @Wilczek2014 evaluate the concordance between outline-based approaches, the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. They found that the variation in ceramic shapes analyzed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes. In addition, the results of shape variation by EFA helps to understand the level of production standardization over time across that region.
<--Topi2007-->

@@

# Materials and archaeological background

@@

## social changes in cross-cultural interaction contexts
- cultural resistance
- trade [@Alizadeh2018]
- identity, social boundary, Raux2015, Boness2017

The results confirm that the differences in pottery shapes can be detected using outline analysis approaches. It shows there is differences in pottery shapes between pre-European and post-European, and Pre-European and Chinese contact. An increasing standardization of shape of pottery during the Chinese contact period compared to the other two periods is observed. This study provide an comparable basis for our understanding of the standardized pottery as an indicator of social changes in a cultural contact conditions. Northeastern Taiwan had been encountered two major foreign contacts in the 17th century with the Europeans and 19th century with the Chines immigrants. It shows that the standardization of pottery shape appears during the European presence and increased in the Chinese period. This may suggest that an emergence of specialized groups or a consensus of the adoption of similar shape of pottery in the society. When it comes to ceramic specialization, metric measurements and clay paste are two major indicators. Shape analysis could be viewed as a replacement of metric measurement that enables subtle changes to be detected. We observe the increasing standardization of the pottery shape, however, it is interesting to note that there is no obvious changes in the clay paste across different phases. In the vitiation that the shape become more standardized without changes in raw materials,
